---
title: "The Effect of Ambient Pressure on the Response of FID and TCD"
subtitle: "Technical Note"
description: |
    Technical Note
output:
    rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Technical Note}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
header-includes:
  - \usepackage[subrefformat=simple]{subfig}
---

```{r setup, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Summary

Presented here is a package developed for **gas chromatography** which models a 
series of measurements of control samples with the objectives of retrospectively 

* correcting the effect that atmospheric pressure has on the response of an FID 
or a TCD,
* compensating instrumental bias and drift, and
* linearizing the detector response

at the same time.  

The calculation outputs two objects:

* a correction object containing a pressure correction coefficient which is a single 
parameter depending only on atmospheric pressure and 
* a drift object containing a number of daily drift factors which depend on the 
peak areas observed.

Since the artifact generated by atmospheric pressure is diluted by instrumental 
artifacts such as baseline shift or drift which cause bias in the data, the code 
negotiates the experimental errors between the drift object and the correction object. 
This vignette describes the theoretical background of the experimental errors and 
gives insight into the algorithm deployed.

Applying this algorithm to experimental data has shown that the repeatability
could be reduced to 0.1% - 0.2%.

## Glossary

### Abbreviations
* EPC: electronic pressure control; electronic pneumatic controller
* FID: flame ionization detector
* GC:  gas chromatograph; gas chromatography
* S/SL: split/splitless; type of a GC injector
* TCD: thermal condctivity detector

### Symbols
$Y$:  measured analyzer output (e.g. peak area, amount, peak height); here referred to as peak area  
$\widehat{Y}$:  true value of peak area  
$Y^\#$:  corrected peak area  
$P$:  measured atmospheric pressure  
$dY/dP$: slope of the fit of $Y \sim P$  
$P_{ref}$:  reference pressure  
$Y_{ref}(episode)$: the expected peak area at reference pressure of an episode  
$\overline{Y}_{corr}$: the mean of the pressure corrected areas of the full data set  
$\kappa$:  atmospheric pressure correction coefficient  
$\delta(Y, d)$:  drift factor  
$\epsilon$:  statistical error  
$M$:  pressure compensation coefficient according to Ayers  
$r_{br}$:  branching ratio; total ion formation rate divided by the overall reaction (burning) rate  
$\theta$:  adiabatic flame temperature  
$x_C$:  mole fraction of carbon in the feed gas  
$F_d$:  detector feed gas flow rate (fuel gas + air + makeup gas + column effluent)  
$M_G$:  average molar mass of the detector feed gas  
$M_C$:  molar mass of carbon  
$P_d$:  atmospheric (= detector) pressure  
$R$:  gas constant  
$T$:  temperature of the feed gas  
$\dot{n}$:  ion formation rate  
$\eta_c$: charge collection efficiency  
$V$: output voltage of the detector  
$I$:  filament current  
$T$:  filament temperature  
$T_w$:  block temperature  
$R_0$:  filament resistance at 0°C  
$\alpha$:  temperature coefficient of the sample  
$x$:  molar fraction of the sample  
$\lambda_1$:  thermal conductivity of carrier gas  
$\lambda_2$:  thermal conductivity of sample  
$A, B$:  constants  

## Introduction

The effect that ambient pressure has on an FID signal was first reported by Bocek et al.
in 1969 [1]. The authors claim that ionization efficiency is strongly dependent on pressure
and that normal changes in atmospheric pressure can result in changes to as much
as ±5% of the value measured at 1 atm.

US patent 4,512,181 [2] describes a procedure how to compensate the impact that atmospheric 
pressure variations have on the output of a gas chromatographic analyzer. They introduce a compensation 
coefficient $M$ which should be determined in conjunction with the calibration of the instrument. The authors 
quantificate the pressure effect as

$$ \frac {Y} {\widehat{Y}} = 1 + M \left( \frac{P}{P_{ref}} - 1 \right) \tag{1}$$
where  
$P$:  atmospheric pressure at the time the sample was measured  
$P_{ref}$:  atmospheric pressure at calibration time  
$Y$:  measured analyzer output (e.g. peak area, amount, concentration) at pressure $P_a$  
$\widehat{Y}$:  expected analyzer output at $P_{ref}$  
$M$:    pressure compensation coefficient; the slope of plot $\frac{Y}{Y^\#}$ as 
a function of $\frac{P}{P_{ref}}$

and the corresponding compensation function of the detector response (peak area or peak height) of a peak as 

$$ Y^\# = \frac{Y}{1 + M \left( \frac{P}{P_{ref}} - 1 \right)} \tag{2}$$
where  
$Y^\#$:  analyzer output as if the sample had been measured at $P_{ref}$  

The pressure artifact had caught the attention of an instrument manufacturer who 
invented electronic pressure control (EPC), which tries to remedy the problem by 
measuring atmospheric pressure and stabilizing pressures and flows of an instrument accordingly
[3]. Despite these efforts the manufacturer still specifies an area precision at 
better than 1% RSD [4]. This figure is in accordance with the long-run variances 
obtained by our experiments without applying any correction or compensation.

The pressure artifact seems to have been largely overlooked in the literature since then. 
Barwick does not mention atmospheric pressure as a source of uncertainty in flame 
ionization detection in her excellent and comprehensive review [5]. Nor does this 
topic appear in a standard for the testing of flame ionization detectors [6]. 
A pressure influence on FID and TCD response is even expressly rejected in another 
standard [7].

## Instrumental details

### Theory of the flame ionization detector

Theory is that a flame ionization detector is a mass sensitive detector. An atmospheric 
hydrogen-air flame forms its core. Any organic compound that enters the flame produces ions
in an amount which is almost perfectly proportional to the mass of carbon atoms reaching the flame [8].

Technically speaking, the FID measures the rate at which these ions are produced 
and subsequently collected. The output voltage $V$ of the detector is expected to be proportional 
to the ion formation rate $\dot{n}$ and the charge collection efficiency $\eta_c$.

$$V \propto \eta_c \cdot \dot{n} = \eta_c(P, \theta)   \left[ r_{br}(\theta) \cdot \frac{x_CM_C}{M_G} \right] \left[ F_d \cdot \frac{P M_G}{RT} \right] \tag{3}$$

where  
$r_{br}$:  branching ratio; total ion formation rate divided by the overall reaction (burning) rate  
$\theta$:  adiabatic flame temperature  
$x_C$:  mole fraction of carbon in the feed gas  
$F_d$:  detector feed gas flow rate (fuel gas + air + makeup gas + column effluent)  
$M_G$:  average molar mass of the detector feed gas  
$M_C$:  molar mass of carbon  
$P_d$:  atmospheric (= detector) pressure  
$R$:  gas constant  
$T$:  temperature of the feed gas

The expression in the left square bracket gives the mass ratio of ions formed in the burning 
gas, while the expression in the right square bracket represents the mass flow rate into
the flame. The equation suggests that the mass flow rate into the detector is proportional 
to pressure; however, this only holds true if the gas flows to the detector are not controlled by an EPC.

Ions and electrons are produced in the flame in equal amounts. An inhomogeneous electrostatic field 
spreads between collector and jet. The collector is a tube which is aligned at a 
small distance above the jet, see Fig. 1. This collector is negatively charged with respect 
to the jet and thus collects the positively charged ions, while the jet collects 
the electrons. The movement of charge produces a small current which is amplified and converted to 
a voltage in the detector electronics. Ions and electrons that have not been captured 
recombine so that there remains a net zero charge. 

It is well known that an FID will saturate at high analyte loading. This saturation 
is caused by a space charge effect. When a massive amount of ions is generated, 
these ions screen the electrostatic field to some extent and thus slow down the migration speed 
of the ions. Consequently, ions and electrons have more time to recombine and the 
sensitivity of the detector decreases.

Pressure also affects flame geometry, but in a complicated way. Flame speed $S_L$, i.e. the 
rate at which the flame front expands, decreases with increasing pressure [8a]. 

$$ S_L (cm/s) = 43[P (atm)]^{-0.5} $$
At the same time flame stretch, which measures the rate at which the area of a flame 
surface grows, increases. In simple words: As the pressure increases, flame height 
increases and flame width decreases. 

Empirical relation of Flame length $L_f$ of turbulent jet non-premixed flames

$$ L_f = \frac{23 \cdot d_j \cdot \sqrt{\frac{\rho_B}{\rho_L} } }{Y_{B,st}} $$

When the flame height increases, the flame starts to be partly located inside the collector tube,
where the electrostatic field strength and the local charge collection efficiency 
are zero, cf. Fig. 1.

Our measurements indicate that FID response typically changes at a rate between -0.7% 
and -0.8% per 10 hPa per 10 hPa of pressure increase. (A rate of -1.0% would indicate that 
detector response and pressure are inversely proportional.)


![Fig. 1. Cutaway diagram of an FID (Source: www.agilent.com)][Fig 1.]{width=100%}

> **Note:**  
Only instruments which are equipped with an EPC show a distinctive negative rate 
of the pressure effect. Flame length is directly proportional to the volumetric 
flowrate [8a]. Given a constant mass flow rate, volumetric flow rate is inversely 
proportional to pressure. An EPC in constant flow mode provides a constant mass 
flow rate.  
Instruments having manual pneumatics [8c], provide a constant volumetric flow rate 
independent of pressure, and so is the flame length.


<span style="color:red">
**Question:**  
The Agilent manuals provide only scarce information on what an EPC exactly does.
I have no access to a GC anymore. Can someone please check the above note 
experimentally? Thank you.
</span>

### Theory of the thermal conductivity detector

Theory says that the a thermal conductivity detector is a concentration sensitive detector. 
The TCD has its application in gas analytics and process analytics where it is used for the detection 
of permanent gases and light hydrocarbons.

According to van Es et al. [9] the output voltage $V$ of a TCD is given as

$$ V = I \cdot R_0 \cdot \alpha \cdot x 
       \left[ A - \frac{\lambda_1}{\lambda_2 B} \right] 
       \left( T - T_w \right) 
       \frac{1 + \alpha T}{1 + \alpha T_w} \tag{4}$$
where  
$I$:  filament current  
$T$:  filament temperature  
$T_w$:  block temperature  
$R_0$:  filament resistance at 0°C  
$\alpha$:  temperature coefficient of the sample  
$x$:  molar fraction of the sample  
$\lambda_1$:  thermal conductivity of carrier gas  
$\lambda_2$:  thermal conductivity of sample  
$A, B$:  constants  

Van Es et al. [9] conclude: "The thermal conductivity of carrier gas and component 
are independent of pressure at least down to 0.1 bar. Consequently the TCD sensitivity 
increases inversely proportional with pressure."

Our measurements, however, show that the sensitivity of a TCD typically changes 
at a rate of about -0.4% to -0.5% per 10hPa of pressure increase, which is only 
half of the projected rate.

![Fig. 2. Cutaway drawing of the Agilent single cell TCD (Source: www.agilent.com)][Fig 2.]{width=100%}


### Pressure effects during sample introduction

Pressure is known to affect the response of an instrument when gaseous samples are 
concerned: In this case the sample amount is proportional to the pressure inside the sample 
loop. Accordingly this applies to the amount of the analyte and the measured peak area. 

![Fig. 3. Schematic of a gas sampling valve (Source: www.vici.com)][Fig 3.]{width=100%}

Please note that pressure has a positive effect on sample amount while its effect 
on detector response is a negative one. For an ideal gas the rate of response change per pressure 
change is +1.0% per 10hPa pressure increase. So injecting a gaseous sample at atmospheric 
pressure counteracts the effect that atmospheric pressure has on the respective 
detector. 

### Electronic pressure / flow control

Electronic pneumatic controllers [3, 10] have been replacing manual pneumatic modules since
the mid-1990s. EPC modules are easy to operate and guarantee a stable gas flow resp. 
column head pressure regardless of
atmospheric pressure. The prime objective of its inventors was to achieve better retention
time stability [3]. Area precision was not in their focus.

Fig. 4 illustrates the workings of EPC on an S/SL inlet in split mode. Please follow 
the explanations given in [10]. During a split injection, a major part of the sample 
is vented from the inlet to the atmosphere while only a minor part is introduced 
into the column. The split ratio is set by the user in the analytical method. The
firmware of the instrument calculates the setpoints of the pressure sensors in 
order to obtain the required split, septum purge and column flow rates. Due to the
calculation these flows are supposed to be independent of atmospheric pressure and the
split ratio is held constant.

The above also applies to the detector feed gases when controlled by an EPC.

![Fig. 4. Schematic of a Split/Splitless Inlet (Source: www.agilent.com)][Fig 4.]{width=100%}

### Origin of bias

Bias is characterized by the fact that not all measurements in a series belong to 
the same population, but that several independent populations can be observed over 
time. This can be visually identified in a plot of measured peak areas over time (i.e. 
a local fit). Signs of bias can be a heavy-tailed or skewed distribution of the 
residuals; in extreme cases several modes can appear, see Fig. 1.

```{r show_autocorrelation, echo = TRUE, fig.width=7.2, fig.height=6}
library(appac)
data("packed_FID")
P.ref <- median(packed_FID$air.pressure, na.rm = TRUE)
appac.colnames <- list(
  sample.col    = "sample.name",
  peak.col      = "peak.name",
  date.col      = "injection.date",
  pressure.col  = "air.pressure",
  area.col      = "raw.area"
)
appac.control <- list(min.data.points = 50, drift.model = "quadratic")
x <- suppressWarnings(appac(
  data = packed_FID,
  P.ref = P.ref,
  appac.control = appac.control,
  appac.colnames = appac.colnames
))

s <- "KGM.11D.4"
p <- "C3H8"
colors <- list(
  highlight_color = "black", lowlight_color = "darkgrey",
  line_color = "darkblue", fill_color = "lightblue"
)

suppressWarnings(appac:::plot_local_fit(
  data = x, sample = s,
  peak = p, coefs = coefficients(x),
  colors = colors, show.compensated.areas = FALSE,
  plot.residuals = TRUE, bins = 50
))
```

Fig. 5. Local fit of raw data exhibiting bias.

We observed that switching an instrument off and on again after some time always
causes bias. The author attributes this effect to the fact that miniscule changes 
occur in the geometry of the detector assembly after cooling down and heating up 
the detector, which in return can affect the electrostatic field between collector and 
jet. Thus we recommend to leave the instruments powered-on day and night. Gas 
savers may safely be applied; no detrimental effects of gas savers have been 
observed so far.

Bias causes autocorrelation in the data. The autocorrelation function can be 
calculated by the acf() function of the stats package, see Fig. 6 and 7.

```{r calc_autocorrelation_raw, echo = TRUE, fig.width=7.2, fig.height=3}
acf(rawAreas(x@correction, s)[, p], pl = TRUE)
```

Fig 6. Plot of the autocorrelation function of the raw areas of the above data set.

```{r calc_autocorrelation_corr, echo = TRUE, fig.width=7.2, fig.height=3}
acf(compensatedCorrectedAreas(x@correction, s)[, p], pl = TRUE)
```

Fig. 7. For comparison: Plot of the autocorrelation function of the adjusted areas.

A consequence of autocorrelated data is that the covariance matrix is positive 
definite; the variance estimators are biased and inconsistent. It is essential to 
remove the autocorrelation before a generalized linear model can be meaningfully 
applied.

## Error and correction model

The appac package features the error model

$$ Y = \widehat{Y} \left( 1 + \kappa(P - P_{ref}) \right) \cdot \delta(Y, d) + \epsilon \tag{5}$$
where  
$Y$:  measured peak area  
$\widehat{Y}$:  true value of peak area  
$\kappa$:  atmospheric pressure correction coefficient  
$P$:  measured atmospheric pressure  
$P_{ref}$:  reference pressure  
$\delta(Y, d)$:  drift factor  
$\epsilon$:  statistical error  

The corresponding correction model follows as

$$ Y^\# = \frac{Y} {\delta(Y, d) \cdot \left( 1 + \kappa(P - P_{ref}) \right)}  \tag{6} $$

where  
$Y^\#$:  corrected peak area

Equation (6) resembles equation (2) if the drift factor $\delta(Y, d)$ 
is disregarded. 

$$ Y^\# \approx \frac{Y} { 1 + \kappa(P - P_{ref}) } \tag{7};  \kappa \equiv \frac{M}{P_{ref}} $$

Since we wanted the correction to be independent of a calibration, we applied the 
correction to measurements of control samples. The method presented here differs 
from that proposed by Ayers et al. [2] in that the correction coefficient $\kappa$ is 
universally applicable to each peak in a chromatogram, whereas the compensation 
coefficient $M$ in Ayers' approach is calculated for each peak individually.

To achieve this goal, the pressure dependencies of the peak areas of all individual 
peaks (local fits) are determined. The results of the local fits are then summarized. 
From this summary, the correction coefficient is determined in a global fit, the 
result of which is valid for all peaks of the underlying chromatographic channel. 
This is a three-step process:

* First determine a reference pressure $P_{ref}$ that lies at the centre of the 
data sets concerned.  
* For each peak of each control sample fit peak area $Y$ vs atmospheric pressure $P$ 
with a linear model to obtain the slope $dY/dP$ of the fitted line and the expected 
peak area $Y_{ref}$ at $P_{ref}$. We will call this a local fit.  
* Then fit $dY/dP$ vs $Y_{ref}$ of all local fits to a quadratic model 

$$dY/dP \sim 0 + Y_{ref} + Y_{ref}^2 \tag{8}$$
We call this the global fit. The fitted slope can be identified as $\kappa$. 
The quadratic tern in eq. (8) is only used in the beginning of the calculation. 
Further in the algorithm, a linear fit is sufficient.

However, a fit according to (7) is called a naive fit, because it neglects 
instrumental artifacts such as baseline shift. A naive fit is applicable only to 
short time periods, where bias or trend cannot be observed in the data. These are 
caused by above-mentioned instrumental artifacts and lead to a time-dependent bias 
in the data. This bias in return induces a high degree of autocorrelation in the 
measurements series. The autocorrelation in the data must be removed before an 
attempt can be made to fit a pressure correction for longer periods of time.

An analysis of the available data has shown that autocorrelation does not exist for sufficiently 
short segments of the time series. Autocorrelation is caused by structural breaks 
within the data. Consequently, we will identify the breakpoints at which the structural 
breaks happen and slice the time series accordingly into segments, which we call episodes.

A local fit (of area vs pressure) 

$$Y \sim P \tag{9}$$

like in the naive fit can then be applied to each of the episodes individually. 
A quadratic global fit of 

$$dY/dP \sim 0 + Y_{ref} + Y_{ref}^2 \tag{10a}$$ 

follows which includes all results of the local fits. 

This first global fit, however, is overfitted because the bias underlying the 
episodes has not yet been removed. A mean value of $\overline{Y}_{ref}$ is established
as the mean values of ${Y}_{corr}$: $\overline{Y}_{corr}$ for each acquisition interval (daily, weekly, etc.)

$$\overline{Y}_{ref} \equiv mean(\overline{Y}_{corr}) \tag{11}$$

With $\overline{Y}_{ref}$, $bias(episode)$ is quantified for each episode as 

$$bias(episode) = Y_{ref}(episode) - \overline{Y}_{ref} \tag{12}$$
where

$Y_{ref}(episode)$: the expected area at reference pressure of an episode  
$\overline{Y}_{corr}$: the mean of the pressure corrected areas of the full data set

The calculated bias is subtracted from the measured peak areas to obtain debiased
raw areas, which are further corrected according to eqn. (7). This returns bias 
compensated and pressure corrected peak areas $Y_{debias, corr}$. 

Now a first estimate of the drift can be calculated. Daily drift factors $\delta(Y, d)$ 
are obtained by fitting
$$ Y_{debias, corr}  \sim  \overline{Y}_{ref} (+ \overline{Y}_{ref}^2) \tag{13}$$
to all peaks measured on the same day. The fitted polynomial coefficients are 
$b_0$, $b_1$ and $b_2$. Accordingly, the daily drift coefficient $\delta(Y, d)$ 
is calculated as
$$ \delta(Y, d) \approx b_0  + b_1 Y \left( + b_2 Y^2  \right) \tag{14} $$

<span style="color:red">
**Question:**  
How can this approximation be avoided?
</span>

> **Note:**  
Although it is possible to calculate the drift for a single control sample, the value 
thus obtained is not menaningful for predicting the drift of other samples.

In the next step the drift coefficients are applied to the raw peak areas, and compensated
raw areas $Y_{raw,comp}$ are obtained. 

Now $Y_{raw,comp}$ is subjected to a pressure fit according to eq. (9) and 

$$dY/dP \sim 0 + Y_{ref} \tag{10b}$$ 

From this stage on, $\kappa$ can be fitted using a linear fit as the p-values show
that the formerly essential quadratic coefficient is insignificant. Using the 
newly obtained value of $\kappa$, the raw areas are corrected according 
to eq. (7) and then drift is recalculated according to eq. (13) and (14). Repeatedly, 
corrected and compensated corrected are calculated with raw areas as the input.

The value of $\kappa$ has been improved after this round; it is time now to polish
the drift. This requires that any remaining bias in the residual is removed.

We do this in the following way: First we calculate expected and residual areas with 
the current drift factors and correction coefficient. Next we fit the residuals 
against pressure to remove any hints of uncompensated pressure dependence from 
the residual areas

$$ r_Y \sim P; r_Y = Y - \widehat{Y} \tag{15} $$

In this way, correction values $corr(dY/dP)$ for the slopes $dY/dP$ and the reference areas $Y_{ref}$: $corr(Y_{ref})$
of the local fits are obtained, see eq. (8). Subsequently, the corrections are applied to the original values and 
corrected values of the slopes and reference areas are obtained.

$$ {dY/dP}_{corr} = dY/dP + corr(dY/dP) \tag{16} $$

$$ Y_{ref, corr} = Y_{ref} + corr(Y_{ref}) \tag{17} $$

Using these corrected slopes and reference areas, a new global fit, eq. (10b) is 
triggered. Following, corrected areas, drift, compensate, expected and residual 
areas are recalculated. 

At this point, the final results of drift and correction are achieved.

<span style="color:red">
**Question:**  
There is still a small bias in $Y_{ref}$. The mean of the residuals is not exactly 0. 
Should this bias be removed?
</span>

## Acknowledgements

The author would like to sincerely thank Werner Hässelbarth, formerly of BAM, for 
proof-reading the manuscript.

## References

[1] Bocek, P., Novak, J., Janak, J., Effect of Pressure on the Performance of the 
Flame Ionization Detector, J. Chromatogr. 43, 431-436 (1969)

[2] Ayers, B. O., Clardy, E. K., Pressure Compensation for a Chromatograph, 
US Patent 4,512,181 (1985)

[3] Henderson, R., Wikfors, E., Retention time stability in a gas chromatographic
apparatus, US Pentent 5,476,000 (1995)

[4] The importance of area and retention time precision in gas chromatography, Technical
Note, Agilent Technologies (2005)

[5] Barwick, V. J., Sources of uncertainty in gas chromatography and high-performance 
liquid chromatograph, J. Chromatogr. A , 849, 13-33 (1999)

[6] ASTM Standard E594-96 (2019), Standard practice for testing flame ionization detectors 
used in gas or supercritical fluid chromatography, ASTM International

[7] ISO standard 6974-1:2012, Natural gas — Determination of composition and 
associated uncertainty by gas chromatography Part 1: General guidelines and 
calculation of composition, International Organization for Standardization

[8] Schofield, K., The enigmatic mechanism of the flame ionization detector: Its 
overlooked implications for fossil fuel combustion modeling, Progress in energy 
and combustion science 34, 330-350 (2008)

[8a] Turns, S. R:, An introduction to combustion, Concepts and applications, McGraw-Hill,
Boston (2011)

[8c] Vici Valco Instruments, Model 100 flow controller, Technical Note 613, Vici AG International (2011)

[9] van Es, A. J. J., Cramers, C. A. M. G., & Rijks, J. A., Detection limits of thermal conductivity and photoionization detectors in high speed narrow bore CGC, Journal of High Resolution Chromatography 12(5), 303-307 (1989)

[10a] Taylor, T., Electronic flow and pressure control in gas chromatography: Opening 
the lid of the black box!, LCGC international 17(11), 24-28 (2021), <https://www.chromatographyonline.com/view/electronic-flow-and-pressure-control-in-gas-chromatography-opening-the-lid-of-the-black-box-

[10b] Hinshaw, J. V., The electronic control of carrier gas pressure, flow, and 
velocity, LCGC North America 32(10), 786-795 (2014). https://www.chromatographyonline.com/view/electronic-control-carrier-gas-pressure-flow-and-velocity

[Fig 1.]: https://www.agilent.com/cs/publishingimages/FID-HIW-Image-960x540.jpg

[Fig 2.]: https://www.agilent.com/cs/publishingimages/TCD-HIW-Image-960x540.jpg

[Fig 3.]: https://www.vici.com/support/images/app11.gif

[Fig 4.]: https://www.agilent.com/cs/publishingimages/split-mode-flow-diagram-960x540.jpg
